Simulating High-Dimensional Multivariate Data using the bigsimr R Package

### 2020-10-22 050-examples.Rmd

# Example applications for our motivating data

@@

```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
## full workflow simulating RNA-seq data
brca <- readRDS(file = 'data/brca99.rds') 
```

@@

```{r ch050-estMargins, echo = FALSE, eval = TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    if (p < minP) {p <- minP } ## maybe add noise here
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    return( list("nbinom", size = n, prob = p) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## describe the margins
probs <- unlist( lapply( brcaMargins, function(x) {x$prob} ) )
## summary(probs)
sizes <- unlist( lapply( brcaMargins, function(x) {x$size} ) )
## summary(sizes)
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

@@

## UHD Parameteric bootstrap

Since many of the correlations are near zero, one could ask whether it is
possible that test the hypothesis that $H_0: R = R_0$.

```{r ch050-corrTest, echo=FALSE, eval=FALSE, include=FALSE}

```

>>>>

# Example applications for our motivating data {#examples}

@@

```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
## full workflow simulating RNA-seq data
brca <- readRDS(file = 'data/brca99.rds')
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

@@

```{r ch050-estMargins, echo = FALSE, eval = TRUE}
## 2. Estimate NegBin parameters using Method of Moments
## estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
estimateNegBinMoM <- function(tmpGene) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    ## if (p < minP) {p <- minP } ## maybe add noise here
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    return( list("nbinom", size = n, prob = p) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## describe the margins
probs <- unlist( lapply( brcaMargins, function(x) {x$prob} ) )
## summary(probs)
sizes <- unlist( lapply( brcaMargins, function(x) {x$size} ) )
## summary(sizes)
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

Here we use the basic method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
We model marginals distributions as coming from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.

TYPESTE MoM estimators here.

@@

## Monte Carlo estimation of MSE

Monte Carlo (MC) methods offer flexible and straightfoward statistical inferenec in this typically difficult analytic setting.
MC methods are routinely used many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
For an concise introduction to these Methos, see, for example @Rizzo2007, Ch. 6.
As a simple example, let's now employ `bigsimr` to estimate the mean squared error (MSE) of the method of moments estimators we used in section X.

Specifically, 
TYPESET THE MATH HERE. REF to MoM above.

```{r ch050-mom-mse, echo=TRUE, eval=TRUE, cache=FALSE}
## system( "rm results/mseSizes.rds")
if ( !file.exists( 'results/mseSizes.rds') )  {
    n <- nrow(brca)
    m <- 100
    momSizes <- replicate(n = m, expr =
                                     { tmpSim <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = cores)
                                         tmpMom <- apply( unname(as.matrix(tmpSim)), 2, estimateNegBinMoM )
                                         tmpSizes <- unlist( lapply( X = tmpMom, function(x) x$size ) )
                                     }
                          )
    sqDiffSizes <- apply( momSizes, 2, function(x) { (x  - sizes)^2 } )
    mseSizes <- rowMeans( sqDiffSizes )
    saveRDS(mseSizes, 'results/mseSizes.rds')
}
```

```{r ch050-fig-mseSizes, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
mseSizes <- readRDS(file = 'results/mseSizes.rds')
qplot(mseSizes, bins = 20) + theme_bw()
```

One could then compare the MSE of the baisc MoM estimation scheme to other strategies ( such as MLE, trimmed approaches, etc.) and decide whether a more computational costly approach is warrented or if the MoM will suffice.
